---
title: "A Codebook for the run_analysis.R script for getting and cleaning data project"
author: "Alemu Tadesse"
date: "8/11/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## The task that is accomplished with this script (run_analysis.R) are the following 

* Merge the training and the test sets to create one data set.
* Extract only the measurements on the mean and standard deviation for each measurement. 
* Use descriptive activity names to name the activities in the data set
* Appropriately label the data set with descriptive variable names. 
* Create a second, independent tidy data set with the average of each variable for each activity and each subject

### The first step is to download the data 

To download data files for processing from https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip  
(by just clicking and placing the zip file)

## Reading the files 

```{r,message=FALSE,echo=FALSE}
## Load required libraries
library(data.table)
library(dplyr)
library(reshape2)

Activity <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/activity_labels.txt")
features <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/features.txt")
X_test <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/test/X_test.txt")
Y_test <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/test/y_test.txt")
subject_test <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/test/subject_test.txt")
activity_labels <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/activity_labels.txt")
X_train <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/train/X_train.txt")
Y_train <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/train/y_train.txt")
subject_train <- read.table("C:/Users/antoe76/Downloads/getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/train/subject_train.txt")
```

```{r,message=FALSE}
#Create activity labels
Activity
Activity_labels<-Activity[,2] ### Extract the labels 
# Load: featires

#Create features labels
features<-features[,2]

head(features)
# Extract features of the mean and standard deviation for each measurement.
extract_features <- grepl("mean|std", features) # Logical data to indicate locations of mean or std features 
head(extract_features)


# Extracting the measurement on the mean and standard deviation. 
#X_test = X_test[,extract_features]
dim(X_test)
#[1] 2947  561
length(features)
#names(X_test) = features
names(X_test) = features
X_test = X_test[,extract_features] # subset X_test satisfying the condiction extract_features


activity_id<-Y_test[,1]

activity_labels<-activity_labels[,2]
Y_test[,2] = activity_labels[activity_id]
names(Y_test) = c("ActivityID", "ActivityLabel")
#View(subject_test) # check what the dat looks like 
names(subject_test) = "Subject"

# since subject_text, x_test and y_test have the same number of rows we can combine them columnwise (cbind)
test_data <- cbind(subject_test, Y_test, X_test)
#names(test_data)<-c("Subject","ActivityID","ActivityLabel")
# Load the X_train & y_train data and apply a similar manipulation as the test data sets shown above 

names(X_train) = features
# Extract only the measurements on the mean and standard deviation for each measurement.
X_train = X_train[,extract_features]


train_activity_id<-Y_train[,1]
Y_train[,2] = activity_labels[train_activity_id]

# Load activity data

names(Y_train) = c("ActivityID", "ActivityLabel")
names(subject_train) = "Subject"

# Combine the training dataset
train_data <- cbind(subject_train, Y_train, X_train)

# rename the first ciluymn of test_data as "Subject" - Upper case "S" and then combine the datasets
names(test_data)[names(test_data) == 'subject'] <- 'Subject'
test_and_train_data = rbind(test_data, train_data)
names(test_and_train_data) # names of the combined data set
```
## Relabeling the combined dataset and calculating the average for each subject
```{r,message=FALSE,echo=FALSE}
names(test_and_train_data)<-gsub("Gyro", "Gyroscope", names(test_and_train_data))
names(test_and_train_data)<-gsub("^t", "time", names(test_and_train_data))
names(test_and_train_data)<-gsub("Mag", "Magnitude", names(test_and_train_data))
names(test_and_train_data)<-gsub("^f", "frequency", names(test_and_train_data))
names(test_and_train_data)<-gsub("Acc", "Accelerometer", names(test_and_train_data))
names(test_and_train_data)<-gsub("BodyBody", "Body", names(test_and_train_data))
names(test_and_train_data)<-gsub("tBody", "TimeBody", names(test_and_train_data))
names(test_and_train_data)<-gsub("-mean()", "Mean", names(test_and_train_data))
names(test_and_train_data)<-gsub("-std()", "STD", names(test_and_train_data))
names(test_and_train_data)<-gsub("-freq()", "Frequency", names(test_and_train_data))
names(test_and_train_data)<-gsub("angle", "Angle", names(test_and_train_data))
names(test_and_train_data)<-gsub("gravity", "Gravity", names(test_and_train_data))
final_data <- melt(test_and_train_data, id = c('Subject', 'ActivityID','ActivityLabel'))
final_average_value <- dcast(final_data, Subject + ActivityID ~ variable, mean)

```
```{r,message=FALSE,echo=TRUE}

head(final_average_value)
```